\paperTitle

### 2014-05-17 doc/icml2014/background.tex
intermediate sync

Next, we assume the graphical model only has conditional independences given by
the graph:
\begin{assumption}[Faithful]
  \label{prop:ci}  
  For any hidden variables $a,b,c \in \bh$
  such that there is an active trail between $a$ and $b$ conditioned on $c$,
  we have that $a$ and $b$ are dependent given $c$ (i.e., $a\not\perp b \given c$).
\end{assumption}

@@

\begin{definition}[Bottleneck]
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$
  that are conditionally independent given $h_i$,
  and each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is a bottleneck
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$;
$x_2^a,x_2^b,x_1^b$ are view of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is a bottleneck.  Note that views are allowed to overlap.

@@

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  We have that $\mOpphat{v}{i}$
  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on the $k$-th singular value of
  $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
However, $\sigma_{k}(\mOpp{v}{i})$ can become extremely
small if $h_i$ and $x_v$ are connected via many intermediate hidden variables.

>>>>

Next, we assume the graphical model only has conditional independences given by
the graph:
\begin{assumption}[Faithful]
  \label{prop:ci}  
  For any hidden variables $a,b,c \in \bh$
  such that there is an active trail\footnote{See
  \citet{koller2009probabilistic} for a definition.
  We do condition on observed variables.} between $a$ and $b$ conditioned on $c$,
  we have that $a$ and $b$ are dependent given $c$ (i.e., $a\not\perp b \given c$).
\end{assumption}

@@

\begin{definition}[Bottleneck]
  \label{def:bottleneck}
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$
  that are conditionally independent given $h_i$,
  and each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is a bottleneck
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$;
$x_2^a,x_2^b,x_1^b$ are view of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is a bottleneck.  Note that views are allowed to overlap.

@@

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  We have that $\mOpphat{v}{i}$
  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on the $k$-th singular value of
  $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
However, $\sigma_{k}(\mOpp{v}{i})$ can become extremely
small if $h_i$ and $x_v$ are connected via many intermediate hidden
variables.\footnote{To see this, suppose $h_1$ has a view $x_v$ via a chain:
$h_1 - h_2 \cdots - h_t - x_v$. In this example, if
$\sigma_k(\Pr(h_{i+1} \given h_{i})) = a_k$ for each $i = 1,
\cdots, t-1$, then $\sigma_k(\mOpp{v}{1}) = a_k^t \sigma_k(\mOpp{v}{t})$.}.

### 2014-05-17 doc/icml2014/directed.tex
intermediate sync

The intuitions of the directed grid generalize readily to general graphs.
The key property required for our algorithm to work is as follows:
\begin{property}[Exclusive views]
  \label{prop:exclusive-views}
Given a graphical model $\sG$,
  a subset of hidden variables $S \subset \bh$ is said to satisfy the \emph{exclusive views property} if for
  each $h_i \in S$, the two conditions hold:
  (i) there exists some observed variable
  $x_{v}$ which is conditionally independent of the others $S \backslash \{ h_i \}$ given $h_i$,
  and (ii) the conditional moment matrix $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ have full column rank $k$ and can be recovered.
We call $x_v$ an exclusive view of $h_i$ in $S$.
%If every hidden clique has the exclusive views property, then we say that $\sG$ does as well.
\end{property}

@@

We now show that if a subset of hidden variables $S$ has the exclusive views property,
then we can recover the marginal distribution $\Pr(\bh_S)$.
Consider any $S = \{i_1, \ldots, i_m\}$ with the exclusive views property. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $S$ and define $\sV
  = \{v_1, \ldots, v_m\}$. % be a set of exclusive views for the clique $S$.
By the exclusive views property,
the marginal over the observed variables $\Pr(\bx_\sV)$
factorizes according to the marginal over the hidden variables $\Pr(\bh_S)$
times the conditional moments:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh_S} \Pr(\Sh{S}) 
                    \Pr(x_{v_1} | h_{i_1}) \cdots \Pr(x_{v_m} | h_{i_m}) \\
  &= Z_{S}(\mOpp{v_1}{i_1},\dots,\mOpp{v_m}{i_m}) \\
  &= Z_{S}(\mOppAll),
\end{align*}
where $\mOppAll = \mOpp{v_1}{i_1} \otimes \cdots \otimes \mOpp{v_m}{i_m}$ is the tensor product of
all the conditional moments.
Vectorizing, we have that
$Z_S \in \Re^{k^m}$,
$M_\sV \in \Re^{d^m}$,
and $\mOppAll \in \Re^{d^m \times k^m}$.
Since each $\mOpp{v}{i}$ has full column rank $k$,
$\mOppAll$ has full column rank $k^m$.
Succinctly, $\mO_\sV$ (which can be estimated directly from data)
is a linear function of $Z_S$ (what we seek to recover).
We can solve for the hidden marginals $Z_S$ simply by projecting $\mO_\sV$ onto the column
space of $\mOppAll$:
\begin{align*}
  Z_{S} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}

@@

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse version)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden subset $S$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution $Z_S$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \dots, x_{v_m}\}$.
      \STATE Return $\mH_S \gets \mO_{\sV}( \mOpp{v_1}{i_i}\pinv, \dots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}

@@

We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}).
  But in a general graph, which cliques have this property?
  To provide intuition, we will provide more interpretable sufficient conditions
  and examples.

It is clear that having uniform bottlenecks implies part (ii)
of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
but we will show that part (i) follows as well.

The exclusive views property 
Next, we define a (in analogy with biconnected component)
\begin{definition}[Bidependent set]
Given a graphical model $\sG$, we say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Intuitively
Note that all cliques are bidependent, but 
Bottlenecks provide three views, but to satisfy condition (ii) of \propertyref{exclusive-views},
only two is required.

Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Uniformly bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
% PL: simplify to cliques (this is also bad notation, since \sG is a set of cliques, not nodes)
%Let $\sG' = \{ h_{i_1}, \dots, h_{i_m} \} \subseteq \sG$ be
  %a connected subgraph of hidden variables.
  %If every hidden variable
  %$h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive views property.
  If a graphical model $\sG$ is faithful and uniformly bottlenecked, then
  every bidependent set in $\sG$ has the exclusive views property (and hence so does $\sG$).
\end{lemma}
\begin{proof}
We prove the contrapositive: suppose there exists a biconnected subgraph of $\sH \subset \sG$ consisting of hidden nodes
such that some $h_0 \in \sH$ does not have an exclusive view.
This means each observed variable $x_i$ in the graph is conditionally dependent on some
$h_i \in \sH \backslash \{h_0\}$ given $h_0$.
Consider any such two $x_1, x_2$.
Conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $\sH$ is biconnected.
Thus, there is an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since $\sG$ is faithful, we have $x_1 \not\perp x_2 \mid h_0$.
As this holds for any $x_1,x_2$, $h_0$ cannot be a bottleneck.
\end{proof}

With this lemma in place, we present the full algorithm, $\LearnMarginals$,
in \algorithmref{directed}.

\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE Graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE Marginals $Z_\sC$ for every clique $\sC \in \sG$

      \FOR{each hidden variable $h_i \in \bh$} 
        \STATE Apply $\TensorFactorize$ to learn conditional moments
        $\mOpp{v}{i}$ for every $v \in \sV_{h_i}$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \ldots, h_{i_m}\} \in \sG$} 
\STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

% DONE: move here because not central to story
\paragraph{Remarks.} We note that \propertyref{bottleneck} can be relaxed if some cliques
  share parameters.
For example, in the directed grid model, we can recover the conditional moments $O$ from
  $x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  Therefore, $h_2, h_3$ and $h_4$
  need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

>>>>

The intuitions of the directed grid generalize readily to general graphs.
The key property required for our algorithm to work is as follows:
\begin{definition}[Exclusive views]
  \label{def:exclusive-views}
Let $S \subseteq \bh$ be a subset of hidden variables.
We say $h_i \in S$ has an exclusive view $x_v$
  if the two conditions hold:
  (i) there exists some observed variable
  $x_{v}$ which is conditionally independent of the others $S \backslash \{ h_i \}$ given $h_i$,
  and (ii) the conditional moment matrix $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ have full column rank $k$ and can be recovered.
We say that $S$ has the \emph{exclusive views property} if every $h_i \in S$ has an exclusive view.
%If every hidden clique has the exclusive views property, then we say that $\sG$ does as well.
\end{definition}

@@

We now show that if a subset of hidden variables $S$ has the exclusive views property,
then we can recover the marginal distribution $\Pr(\bh_S)$.
Consider any $S = \{i_1, \ldots, i_m\}$ with the exclusive views property. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $S$ and define $\sV
  = \{v_1, \ldots, v_m\}$. % be a set of exclusive views for the clique $S$.
By the exclusive views property,
the marginal over the observed variables $\Pr(\bx_\sV)$
factorizes according to the marginal over the hidden variables $\Pr(\bh_S)$
times the conditional moments:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh_S} \Pr(\Sh{S}) 
                    \Pr(x_{v_1} | h_{i_1}) \cdots \Pr(x_{v_m} | h_{i_m}) \\
  &= Z_{S}(\mOpp{v_1}{i_1},\dots,\mOpp{v_m}{i_m}) \\
  &= Z_{S}(\mOppAll),
\end{align*}
where $\mOppAll = \mOpp{v_1}{i_1} \otimes \cdots \otimes \mOpp{v_m}{i_m}$ is the tensor product of
all the conditional moments.
Vectorizing, we have that
$Z_S \in \Re^{k^m}$,
$M_\sV \in \Re^{d^m}$,
and $\mOppAll \in \Re^{d^m \times k^m}$.
Since each $\mOpp{v}{i}$ has full column rank $k$,
the tensor product $\mOppAll$ has full column rank $k^m$.
Succinctly, $\mO_\sV$ (which can be estimated directly from data)
is a linear function of $Z_S$ (what we seek to recover).
We can solve for the hidden marginals $Z_S$ simply by projecting $\mO_\sV$ onto the column
space of $\mOppAll$:
\begin{align*}
  Z_{S} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}

@@

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse version)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden subset $S = \{ h_{i_1}, \dots, h_{i_m} \}$ with exclusive views $\sV = \{ x_{v_1}, \dots, x_{v_m} \}$ and $\{\mOpp{v_1}{i_i}\pinv\}$ % (\definitionref{exclusive-views})
    \ENSURE Marginal distribution $Z_S = \Pr(\bh_S)$.
      %\STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \dots, x_{v_m}\}$.
      \STATE Return $\mH_S \gets \mO_{\sV}( \mOpp{v_1}{i_i}\pinv, \dots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}

@@

We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\definitionref{exclusive-views}).
  But in a general graph, which cliques have this property?
  To provide intuition, we will provide more interpretable sufficient conditions
  and examples.

% Define bidependent
First, define the notion of a bidependent set (in analogy with biconnected component),
in which conditionining on one variable does not break the set apart:
\begin{definition}[Bidependent set]
Given a graphical model $\sG$, we say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent sets can have more conditional independences.
this will be important for pseudolikliehood in \sectionref{pseudolikelihood}.

% bidependent, bottleneck => exclusive views
Bidependent sets are significant because they guarantee exclusive views if they are bottlenecks:
%If a subset of hidden nodes $S$ is a bottleneck, it is clear that (ii)
%of the exclusive views property (\definitionref{exclusive-views}) via \TensorFactorize,
%but we will show that part (i) follows as well.
%Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
  Let $S \subseteq \bh$ be a bidependent subset of hidden variables.
  If $S$ is a bottleneck, then $S$ has the exclusive views property.
\end{lemma}
\begin{proof}
Let $S$ be a bidependent subset and fix any $h_0 \in S$.
Since $h_0$ is a bottleneck, it has three conditionally independent views,
say $x_1, x_2, x_3$ without loss of generality. 
For condition (i), we will show that at least one of the views is conditionally independent
of $S \backslash \{ h_0 \}$ given $h_0$.
For the sake of contradiction, 
suppose that each observed variable $x_i$ is conditionally dependent on some
$h_i \in S \backslash \{h_0\}$ given $h_0$, for $i \in \{1, 2, 3\}$.
Then conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $S$ is biconnected.
Thus, there is an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since the graphical model is faithful by assumption, we have $x_1 \not\perp x_2 \mid h_0$,
contradicting the fact that $x_1$ and $x_2$ are conditionally independent given $h_0$.
To show condition (ii), assume, without loss of generality, that $x_1$ is an exclusive view.
Then we can recover $\mOpp{1}{0} = \Pr(x_1 \mid h_0)$ via $\TensorFactorize$.
\end{proof}

%Note that bottlenecks provide three views, which is important for condition (ii) of
%\definitionref{exclusive-views}, only two is required for condition (i).
With this lemma in place, we present our full algorithm, $\LearnMarginals$,
in \algorithmref{directed}.

\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE Bidependent and bottlenecked subset of hidden nodes $S \subseteq \bh$
    \ENSURE Marginals $Z_S = \Pr(\bh_S)$ % for every bidependent subset of hidden variables $S \in \sG$
    \FOR{each hidden node $h_i \in S$ with exclusive view $x_v$} 
      \STATE Apply $\TensorFactorize$ to learn conditional moments
      $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
    \ENDFOR
    \STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
  \end{algorithmic}
\end{algorithm}

% DONE: move here because not central to story
\paragraph{Remarks.}
Note that even having two independent views for each $h_i \in S$ is sufficient for condition (i)
  of the exclusive views property, while three is needed for condition (ii).
The bottleneck property (\definitionref{bottleneck} can also be relaxed if some cliques
  share parameters.
For example, in the directed grid model, we can recover the conditional moments $O$ from
  $x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  Therefore, $h_2, h_3$ and $h_4$
  need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

### 2014-05-17 doc/icml2014/undirected.tex
intermediate sync

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function $A(\theta)$.
%One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective is a sum over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p^*}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.

>>>>

\subsection{Pseudolikelihood}
\label{sec:pseudolikelihood}

Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function $A(\theta)$.
%One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective is a sum over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p^*}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.
